---
title: "a05"
format: html
author: "Gabriella Chavez"
date: April 01, 2026 
---

## Introduction

This report explores socioecological monitoring data from the Sacramento-San Joaquin Delta. The goal is to visualize patterns in visitor activity across restoration locations and to map where those locations are found. A plot and a map help summarize both the amount of visitation and the spatial distribution of the monitored sites.

## Data

The data for this assignment comes from the Sacramento-San Joaquin Delta Socioecological Monitoring dataset. The file used in this report is `Socioecological_monitoring_data.csv`, saved in the `data/RAW` folder of the project.

```{r}
#| message: false
#| warning: false

### This loads packages needed for reading, plotting, tables, and mapping
library(readr)
library(dplyr)
library(tidyr)
library(forcats)
library(ggplot2)
library(leaflet)
library(DT)
library(scales)
library(janitor)
library(viridis)
library(here)

### This reads the raw dataset from the data folder into R
delta_visits_raw <- read_csv(here("data/RAW/Socioecological_monitoring_data.csv"), show_col_types = FALSE)

### This cleans the column names so they are all lowercase and consistent
delta_visits <- delta_visits_raw %>%
  clean_names()

### This reshapes the data from wide format to long format so visitor types are in one column
visits_long <- delta_visits %>%
  pivot_longer(
    cols = c(sm_boat, med_boat, lrg_boat, bank_angler, scientist, cars),
    names_to = "visitor_type",
    values_to = "quantity"
  ) %>%
  
  ### This renames the location column to make it shorter and easier to use
  rename(restore_loc = eco_restore_approximate_location) %>%
  
  ### This remove the notes column because it is not needed for this analysis
  select(-notes)
```

## Analysis 

```{r}
### This grousp the data by restoration location, date, and visitor type
daily_visits_loc <- visits_long %>%
  group_by(restore_loc, date, visitor_type) %>%
  
  ### This calculates the total number of visits for each location, date, and visitor type
  summarize(daily_visits = sum(quantity), .groups = "drop")

### This groups the summarized data by restoration location
daily_visits_totals <- daily_visits_loc %>%
  group_by(restore_loc) %>%
  
  ### This creates a total number of visits for each restoration location
  mutate(n = sum(daily_visits)) %>%
  
  ### This removes grouping so later steps are easier to control
  ungroup() %>%
  
  ### This reorders the restoration locations by total visits
  mutate(restore_loc = fct_reorder(restore_loc, n))

### This creates a reusable theme object with a white background and larger text
my_theme <- theme_bw(base_size = 14) +
  
  ### This moves the legend to the bottom and remove y-axis tick marks
  theme(
    legend.position = "bottom",
    axis.ticks.y = element_blank()
  )

### This creates a bar plot showing daily visits by restoration location and visitor type
visit_plot <- ggplot(
  data = daily_visits_totals,
  aes(x = daily_visits, y = restore_loc, fill = visitor_type)
) +
  
  ### This draws bars whose lengths represent the number of visits
  geom_col() +
  
  ### This applies a colorblind-friendly color palette
  scale_fill_viridis_d() +
  
  ### This labels the axes, legend, title, and subtitle
  labs(
    x = "Number of Visits",
    y = "Restoration Location",
    fill = "Visitor Type",
    title = "Visits to Delta Restoration Areas by Visitor Type",
    subtitle = "Summed across the study period"
  ) +
  
  ### This sets x-axis tick marks and remove the gap between bars and the axis
  scale_x_continuous(breaks = seq(0, 120, 20), expand = c(0, 0)) +
  
  ### This applies the custom theme to improve readability
  my_theme

### This displays the plot
visit_plot

### This keeps one row for each unique restoration location
locations <- visits_long %>%
  distinct(restore_loc, .keep_all = TRUE) %>%
  
  ### This keep only the columns needed for mapping
  select(restore_loc, latitude, longitude)


### This creates an interactive map centered on the restoration locations
visit_map <- leaflet(locations) %>%
  
  ### This adds a satellite-style base map
  addProviderTiles(providers$Esri.WorldImagery) %>%
  
  ### This adds label tiles on top of the imagery
  addProviderTiles(providers$CartoDB.PositronOnlyLabels) %>%
  
  ### This adds circular markers at each restoration location
  addCircleMarkers(
    lng = ~ longitude,
    lat = ~ latitude,
    popup = ~ restore_loc,
    radius = 5,
    fillColor = "salmon",
    fillOpacity = 1,
    stroke = TRUE,
    weight = 0.5,
    color = "white",
    opacity = 1
  )

### This finally displays the map
visit_map
```

## Discussion

The bar plot shows that visitation differs across restoration locations and also varies by visitor type. The map adds useful spatial context by showing where the monitored locations are distributed across the Delta. Together, these visuals make the data easier to interpret than the raw table one because they show both the amount of visitation and the geographic patterns of the sites. 

```{r}
